Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-01 talk/icooolps2009/paper.tex
Going over some changes by Anto. Still not quite happy with this area.

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}

@@

\title{Tracing the Meta-Level: PyPy's JIT Compiler}

@@

@@

Tracing JITs are an idea explored by the Dynamo project
\cite{bala_dynamo:transparent_2000} in the context of dynamic optimization of
machine code at runtime. The techniques were then successfully applied to Java
VMs \cite{gal_hotpathvm:effective_2006}. It also turned out that they are a
relatively simple way to implement a JIT compiler for a dynamic language
\cite{mason_chang_efficient_2007}. The technique is now
being used by both Mozilla's TraceMonkey JavaScript VM \cite{XXX} and Adobe's
Tamarin ActionScript VM \cite{XXX}.

@@

When a hot loop is identified, the interpreter enters a
special mode, called \emph{tracing mode}. \sout{When in tracing mode, the interpreter
records a history (the \emph{trace}) of all the operations it executes, in addition
to actually performing the operations.
} \anto{During tracing, the interpreter records a history of all the
  operations it executes.}

\anto{
Such a history is called a \emph{trace}: it is a sequential list of
operations, together with their actual operands and results.  By examining the
trace, it is possible to produce highly efficient machine code by emitting
only the operations needed.  Being sequential, the trace represents only one
of the many possible paths through the code: to ensure correctness, the trace
contains a \emph{guard} at every possible point where the path could have
followed another direction, for example \texttt{if}s or indirect/virtual
calls.  When emitting the machine code, we turn every guard into a quick check
to guarantee that the path we are executing is still valid.  If a guard fails,
we immediately quit from the machine code and continue the execution in other
ways.  
}

During tracing, the trace is repeatedly
(XXX make this more precise: when does the check happen?)
checked whether the interpreter is at a position in the program that it had seen
earlier in the trace. If this happens, the trace recorded corresponds to a loop
in the program that the tracing interpreter is running. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be immediately executed,
as it represents exactly the loop that was being interpreted so far.

\anto{XXX I think it's worth spending one more paragraph to explain what a
  trace really is, i.e. that it's a list of \textbf{sequential} operations,
  intermixed to guards which guarantee that this particular sequence is still
  valid.  At the moment, the definition of trace is not given explicitly and
  it's mixed with the details of how the JIT work}

>>>>

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage[utf8]{inputenc}

@@

\title{Tracing the Meta-Level: PyPy's Tracing JIT Compiler}

@@

XXX contributions of this paper include:

@@

Tracing JITs are an idea initially explored by the Dynamo project
\cite{bala_dynamo:transparent_2000} in the context of dynamic optimization of
machine code at runtime. The techniques were then successfully applied to Java
VMs \cite{gal_hotpathvm:effective_2006}. It also turned out that they are a
relatively simple way to implement a JIT compiler for a dynamic language
\cite{mason_chang_efficient_2007}. The technique is now
being used by both Mozilla's TraceMonkey JavaScript VM \cite{XXX} and Adobe's
Tamarin ActionScript VM \cite{XXX}.

@@

When a hot loop is identified, the interpreter enters a
special mode, called \emph{tracing mode}. During tracing, the interpreter
records a history of all the operations it executes.

Such a history is called a \emph{trace}: it is a sequential list of
operations, together with their actual operands and results.  By examining the
trace, it is possible to produce highly efficient machine code by emitting
only the operations needed.  Being sequential, the trace represents only one
of the many possible paths through the code. To ensure correctness, the trace
contains a \emph{guard} at every possible point where the path could have
followed another direction, for example conditions or indirect/virtual
calls.  When emitting the machine code, every guard is turned into a quick check
to guarantee that the path we are executing is still valid.  If a guard fails,
we immediately quit from the machine code and continue the execution by falling
ways.

During tracing, the trace is repeatedly
checked whether the interpreter is at a position in the program that it had seen
earlier in the trace. If this happens, the trace recorded corresponds to a loop
in the program that the tracing interpreter is running. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be immediately executed,
as it represents exactly the loop that was being interpreted so far.
